---
title: "R4DS Ch 27 exercises"
author: "Chris Sirico"
date: "10/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 27.2.1 Exercises

1.  **Create a new notebook using _File > New File > R Notebook_. Read the 
    instructions. Practice running the chunks. Verify that you can modify
    the code, re-run it, and see modified output.**
    
    Did it. Learned Cmd+Opt+I makes a new chunk. Cmd+Return runs the current line of code and Cmd+Shift+Return runs the whole chunk. Neat. Notebooks auto-create html files whether or not you knit to preview. Not so w/ normal RMDs.
    
2.  **Create a new R Markdown document with _File > New File > R Markdown..._
    Knit it by clicking the appropriate button. Knit it by using the 
    appropriate keyboard short cut. Verify that you can modify the
    input and see the output update.**
    
    Cmd+~ switches between preview and editor windows. Along those lines ctrl+tab (mac) switches between documents in the script window. Nice to know.
    
3.  **Compare and contrast the R notebook and R markdown files you created
    above. How are the outputs similar? How are they different? How are
    the inputs similar? How are they different? What happens if you
    copy the YAML header from one to the other?**
    
    Notebook doesn't show the header. Vanilla RMD knit does. output:html_document vs output:html_notebook makes the difference. Changing that field of the YAML header tells R whether to auto-generate html or not.

4.  **Create one new R Markdown document for each of the three built-in
    formats: HTML, PDF and Word. Knit each of the three documents.
    How does the output differ? How does the input differ? (You may need
    to install LaTeX in order to build the PDF output --- RStudio will
    prompt you if this is necessary.)**
    
    Output is predictably in each of the three filetypes.
    
    
```{r}

```

## 27.3.1 Exercises

1.  **Practice what you've learned by creating a brief CV. The title should be
    your name, and you should include headings for (at least) education or
    employment. Each of the sections should include a bulleted list of
    jobs/degrees. Highlight the year in bold.**
    
    # Chris Sirico, Marker-Downer
    ## Education
    - Coursera
    - Liberal arts university, **2005 – 2008**
    - Hard knocks
    
2.  **Using the R Markdown quick reference, figure out how to:**

Add [^1]

[^1]: a footnote.

Add a horizontal rule.

***

> Add a block quote.
    
3.  **Copy and paste the contents of `diamond-sizes.Rmd` from
    <https://github.com/hadley/r4ds/tree/master/rmarkdown> in to a local
    R markdown document. Check that you can run it, then add text after the 
    frequency polygon that describes its most striking features.**
    
    👍
    
## 27.4.7 Exercises

1.  **Add a section that explores how diamond sizes vary by cut, colour, and clarity. Assume you’re writing a report for someone who doesn’t know R, and instead of setting echo = FALSE on each chunk, set a global option.**
```{r}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
plot(diamonds$y, diamonds$cut)
plot(diamonds$y, diamonds$color)
plot(diamonds$y, diamonds$clarity)
```

2.  **Download diamond-sizes.Rmd from https://github.com/hadley/r4ds/tree/master/rmarkdown. Add a section that describes the largest 20 diamonds, including a table that displays their most important attributes.**


```{r setup2, include = FALSE}
library(ggplot2)
library(dplyr)

smaller <- diamonds %>% 
  filter(carat <= 2.5)
```

We have data about `r nrow(diamonds)` diamonds. Only 
`r nrow(diamonds) - nrow(smaller)` are larger than
2.5 carats. The distribution of the remainder is shown
below:

```{r, echo = FALSE}
smaller %>% 
  ggplot(aes(carat)) + 
  geom_freqpoly(binwidth = 0.01)
```

A synopsis of the 20 largest diamonds:

```{r}
largest <- diamonds %>% arrange(desc(diamonds$carat)) %>%
  filter(row_number() <= 20)
summary(largest$y)
```
These babies are about a centimeter in diameter! Daggum.

3.  **Modify diamonds-sizes.Rmd to use comma() to produce nicely formatted output. Also include the percentage of diamonds that are larger than 2.5 carats.**


```{r diamond-sizes-2}
comma <- function(x) format(x, digits = 2, big.mark = ",")

smaller <- diamonds %>% 
  filter(carat <= 2.5)
```

We have data about `r comma(nrow(diamonds))` diamonds. Only 
`r comma(nrow(diamonds) - nrow(smaller))` are larger than
2.5 carats. The distribution of the remainder is shown
below:

```{r, echo = FALSE}
smaller %>% 
  ggplot(aes(carat)) + 
  geom_freqpoly(binwidth = 0.01)
```

`r comma((nrow(diamonds) - nrow(smaller)) / nrow(diamonds)*100)` percent of diamonds are larger than 2.5 carat.

4.  **Set up a network of chunks where d depends on c and b, and both b and c depend on a. Have each chunk print lubridate::now(), set cache = TRUE, then verify your understanding of caching.**

```{r a, cache = TRUE}
lubridate::now()
```

```{r b, cache = TRUE, dependson="a"}
lubridate::now()
```

```{r c, cache = TRUE, dependson="a"}
lubridate::now()
```

```{r d, cache = TRUE, dependson=c("b", "c")}
lubridate::now()
```

Ah, okay. So it'll print the first time it was knit at, starting w/ a and going down the chain, unless cache = FALSE.

