---
title: "R4DS Ch 14-15 Exercises"
author: "Chris Sirico"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(htmlwidgets)
library(forcats) #for gss_cat
```
## 14.2.5 Exercises

@.  **In code that doesn't use stringr, you'll often see `paste()` and `paste0()`.
    What's the difference between the two functions? What stringr function are
    they equivalent to? How do the functions differ in their handling of 
    `NA`?**
    
    paste combines the elements of multiple strings. sep = " " by default, so `paste(c("A","B"), 1, "%")` yeilds the character vector: "A 1 %" "B 1 %".
    the `collapse` parameter instead yields one single string, all smashed together with the assigned character in between: `paste(c("A","B",NA), 1, "%", collapse = " ")` `#> "A 1 % B 1 % NA 1 %"`
    paste0 automatically sets sep = "", a frequent usage case.
    The stringr equivalent is `str_c` (string combine). `str_c` requires character args be wrapped in `str_replace_na` to output literal "NA"s instead of NA values. Base `paste` functions output literal NAs by default. It uses "" as its default separator so it's most comparable to paste0.
    
```{r include=F}
paste(c("A","B",NA), 1, "%", collapse = " ") #A bunch of individual character strings.
paste(c("A","B",NA), 1, "%")
str_c(c("A","B",NA), 1, "%", sep = " ") #A bunch of individual character strings.
str_c(str_replace_na(c(c("A","B",NA), (1), ("%"))),  collapse = " ") #A bunch of individual character strings.
paste0(c("A","B",NA), 1, "%")

paste(1:4, letters[1:4], collapse = "") #2 or more strings pasted element for element.

paste(1:10, collapse = "") #One string smushed together. Here's the sep/collapse rule for each:

# A bunch of individual character strings – You want sep
# 2 or more strings pasted element for element. – You want sep
# One string smushed together.- Smushin requires collapse
# paste0 is short for: paste(x, sep="") So it allows us to be lazier and more efficient.

paste0("a", "b") == paste("a", "b", sep="") ## [1] TRUE
```

@.  **In your own words, describe the difference between the `sep` and `collapse`
    arguments to `str_c()`.**
    
    Collapse outputs one long string separated by the indicated character(s). Sep outputs as many elements as the longest character vector used as an input. Elements from each input are separated by the specified character(s).

@.  **Use `str_length()` and `str_sub()` to extract the middle character from 
    a string. What will you do if the string has an even number of characters?**
    
```{r}
#generate a random letter string, abc
rlength <- abs((rnorm(1)*20%/%1))
abc <- sample((LETTERS), size = rlength, replace = TRUE) %>% str_c(collapse="")

print(c(abc, str_length(coll(abc, locale = "en"))))  # How long is it (collate)? What does it look like?

# make an anonymous function that takes x arg, (feeding in abc after "})" )
(function(x) {

# find the midpoint
midpoint <- as.integer(c(str_length(coll(x, locale = "en"))%/%2, str_length(x)%%2))

# return one- or two-character midpoint
ifelse(midpoint[2] == 0, (str_c(("midletters: "), as.character(str_sub(x, midpoint[1], midpoint[1]+1)))), (str_c(("midletter: "), as.character(str_sub(x, midpoint[1]+1, midpoint[1]+1)))))
})(abc)

```

@.  **What does `str_wrap()` do? When might you want to use it?**
    
    It allows for auto line breaking on long strings. Offers formatting parameters. It would be helpful whenever there's a lot of text for the user of a script, function or program to read in the console.

@.  **What does `str_trim()` do? What's the opposite of `str_trim()`?**
    
    str_trim removes whitespaces characters from the beginning and/or end strings. str_pad adds whitespace.

@.  **Write a function that turns (e.g.) a vector `c("a", "b", "c")` into 
    the string `a, b, and c`. Think carefully about what it should do if
    given a vector of length 0, 1, or 2.**

```{r}
A <- c("a")
AB <- c("a", "b")
ABC <- c("a", "b", "c")
ABCD <- c("a", "b", "c", "d")

# Defining function as nested ifelse statements; hacky, but felt obvious
# What's the more elegant way?
(function(x) {
  # str(x)
ifelse(length(x) == 0, "",  # return nothing for input string length zero
       ifelse(length(x) == 1, x,  # return input for string length one
              ifelse(length(x) == 2, str_c(x[1], " and ", x[2]),   # 2 elements, combine with " and "
                      str_c(str_c(x[1:(length(x)-1)], collapse = ", "), ", and ", str_c(x[length(x)], collapse = ""), collapse = "")  # 3 parts, separate first elements with spaces and then add "and" before last
                     )
              )
       ) %>% return()
})(ABC)

# Results for other inputs:
```

``` {r echo=F}
(function(x) {
  # str(x)
ifelse(length(x) == 0, "",  # return nothing for input string length zero
       ifelse(length(x) == 1, x,  # return input for string length one
              ifelse(length(x) == 2, str_c(x[1], " and ", x[2]),   # 2 elements, combine with " and "
                      str_c(str_c(x[1:(length(x)-1)], collapse = ", "), ", and ", str_c(x[length(x)], collapse = ""), collapse = "")  # 3 parts, separate first elements with spaces and then add "and" before last
                     )
              )
       ) %>% return()
})(A)

(function(x) {
  # str(x)
ifelse(length(x) == 0, "",  # return nothing for input string length zero
       ifelse(length(x) == 1, x,  # return input for string length one
              ifelse(length(x) == 2, str_c(x[1], " and ", x[2]),   # 2 elements, combine with " and "
                      str_c(str_c(x[1:(length(x)-1)], collapse = ", "), ", and ", str_c(x[length(x)], collapse = ""), collapse = "")  # 3 parts, separate first elements with spaces and then add "and" before last
                     )
              )
       ) %>% return()
})(AB)

(function(x) {
  # str(x)
ifelse(length(x) == 0, "",  # return nothing for input string length zero
       ifelse(length(x) == 1, x,  # return input for string length one
              ifelse(length(x) == 2, str_c(x[1], " and ", x[2]),   # 2 elements, combine with " and "
                      str_c(str_c(x[1:(length(x)-1)], collapse = ", "), ", and ", str_c(x[length(x)], collapse = ""), collapse = "")  # 3 parts, separate first elements with spaces and then add "and" before last
                     )
              )
       ) %>% return()
})(ABCD)
# \(^O^)/ woohoo!
```
```{r}
# ♬ Who gives a f*ck about an Oxford comma? ♬
```
## 14.3.1.1 

1.  **Explain why each of these strings don’t match a \: "\", "\\", "\\\".**
  \ is a special character in R and thus must be escaped. "\\" is evaluated from a string into a single backslash into the function it feeds into. R then presumes it to be escaping some other character. "\\\" doesn't work because the first two backslashes become one before the second level of evaluation. The third is evaluated as escaping something else. "\\\\" becomes \\ in the first step, which then escapes the backslash to a single output character.

2.  **How would you match the sequence `"'\`?**

```{r}
(x <- "\"\'\\")
str_view(x, "\"\\'\\\\")
```


3.  **What patterns will the regular expression `\..\..\..` match? 
    How would you represent it as a string?**
    
    It would match a period and any character following it, a second period and any character following it, and a final period and any character following it.
```{r}
(x <- ".a.\t.\ ")
str_view(x, "\\..\\..\\..")
```
  
## 14.3.2.1 Exercises

1.  **How would you match the literal string `"$^$"`?**

```{r}
(x <- "$^$")
str_view(x, "\\$\\^\\$")
```

2.  **Given the corpus of common words in `stringr::words`, create regular
    expressions that find all words that:
    
    1. Start with "y".
    1. End with "x"
    1. Are exactly three letters long. (Don't cheat by using `str_length()`!)
    1. Have seven letters or more.
    
```{r}
str_view(words, "^y", match=T)
str_view(words, "x$", match=T)

# These next two produce long results. Run at your own discretion.
# str_view(words, "^...$", match=T)
# str_view(words, "^.......+", match=T)
```

## 14.3.3.1 Exercises

1.  **Create regular expressions to find all words that:

    1. Start with a vowel.

    1. That only contain consonants. (Hint: thinking about matching 
       "not"-vowels.)

    1. End with `ed`, but not with `eed`.
    
    1. End with `ing` or `ise`.**
    
```{r}
# This result is long.
# str_view(words, "^[aeiou]", match=T) # I'm omitting "y" because it would usually be functioning as a consonant at the beginning of a word.
str_view(words, "^[^aeiouy]+$", match=T) # Here including y because it would be functioning as a vowel in words with no other vowels. I'm a stickler for words. It's my job.
str_view(words, "([^e]+ed$|^ed$)", match=T)
str_view(words, "(ing$|ise$)", match=T)
```

2.  **Empirically verify the rule "i before e except after c".**

```{r}
cies<-(str_match(words, "cie"))
cies<-cies[!is.na(cies)]  

eis<-(str_match(words, "[^c]?hk"))
eis<-eis[!is.na(eis)]

if (length(cies) > 0 || length(eis) > 0) "i is not always before e except after c"

```

3.  **Is "q" always followed by a "u"?**

```{r}
qandnotu <- str_match(words, "q[^u]")
ifelse(length(qandnotu[!is.na(qandnotu)]) != 0, "There's a q without a u!", "No q's without u's here.")
```

4.  **Write a regular expression that matches a word if it's probably written
    in British English, not American English.**
    
```{r}
str_view(words, "[tl]ise$", match=T)
```
    

5.  **Create a regular expression that will match telephone numbers as commonly
    written in your country.**

```{r}
numbs<-c("+39 066 1969084", "704-267-8762", "(704)-267-8762", "1-704-267-8762", "+1 (704) 267 8762", "704.267.8762", "1.704.267.8762", "704 267 8762", ".704.267.8762", "1 704 267 8762", "+1(704)267-8762")

str_view(numbs, "^(^((\\+1)|1)[- \\.]?)?\\(?\\d{3}[\\)-\\.]??\\s?\\d{3}[- \\.]?\\d{4}$", match=T)
```

## 14.4.2 Exercises

1.  **For each of the following challenges, try solving it by using both a single
    regular expression, and a combination of multiple `str_detect()` calls.
    
    1.  Find all words that start or end with `x`.
    
    1.  Find all words that start with a vowel and end with a consonant.
    
    1.  Are there any words that contain at least one of each different
        vowel?**

```{r}
# find words beginning or ending with "x"
str_view(words, "(^x)|(x$)", match=T)
words[!is.na(match(str_detect(words, "x$"), TRUE))|!is.na(match(str_detect(words, "^x"), TRUE))]

# starts with a vowel, ends with a consonant
# str_view(words, "^[aeiou].*[^[aeiouy]]$", match=T) # this produces a long result
sum(str_count(words, "^[aeiou].*[^[aeiouy]]$")) # this verifies 105 results
words[!is.na(match(str_detect(words, "^[aeiou]"), TRUE))&!is.na(match(str_detect(words, "[^[aeiouy]]$"), TRUE))] # %>% length() # this verifies 105 results too

# using look-aheads. Googled this. Assuming the book wants something else.
str_view(words, "(?=.*a)(?=.*e)(?=.*i)(?=.*o)(?=.*u)", match=T) #produces no results for all five vowels

```
2.  **What word has the highest number of vowels? What word has the highest
    proportion of vowels? (Hint: what is the denominator?)**

```{r}
(vowels <- tibble(w = words,
                 vowel = str_count(words, "[aeiou]"),
                 length = str_count(words, "[abcdefghijklmnopqrstuvwxyz]"),
                 vowelp = plyr::round_any(vowel / length, 0.01))) %>% 
  arrange(desc(vowel)) %>% head(10)
  arrange(vowels, desc(vowelp)) %>% head(10) # "a", of course. It's only surprising that "I" didn't make it to the top of the list! ;)
```
##14.4.6.1 Exercises

1.  **Split up a string like `"apples, pears, and bananas"` into individual
    components.**
```{r}
fruit <- "apples, pears, and bananas"
str_split(fruit, boundary("word"), simplify = T) # let's see a matrix

```
    
2.  **Why is it better to split up by `boundary("word")` than `" "`?**

Punctuation and linebreaks can immediately abut words as well as spaces.

3.  **What does splitting with an empty string (`""`) do? Experiment, and
    then read the documentation.**

```{r}
fruit <- "apples, pears, and bananas"
str_split(fruit, "") # equivalent to boundary("character")
# Why do we still lose commas and the word "and"?
```

<!-- ##15.3.1 Exercises # Oops. Extra practice  ¯\_(ツ)_/¯

1.  **Explore the distribution of `rincome` (reported income). What makes the
    default bar chart hard to understand? How could you improve the plot?**
    
    The variables are in an un-intuitive order. We could reorder them:

```{r, fig.width=5,fig.height=3}
incomedist <- gss_cat

ggplot(incomedist)+
  geom_bar(aes(sort(fct_rev(fct_relevel(gss_cat$rincome, "Not applicable", "Refused", "Don't know", "No answer", after = 12))))) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  xlab("Reported Income")
```

1.  **What is the most common `relig` in this survey? What's the most
    common `partyid`?**
    
```{r, fig.width=5,fig.height=3}
ggplot(gss_cat)+
  geom_bar(aes(relig))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) # Protestant

ggplot(gss_cat)+
  geom_bar(aes(partyid))+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) # Finding this line of code very handy for prettier x ticks.
# Independent
```
    
1.  **Which `relig` does `denom` (denomination) apply to? How can you find
    out with a table? How can you find out with a visualisation?**

```{r, fig.width=5, fig.height=3}
# table(gss_cat$relig, gss_cat$denom)
# ^^ This works, although it's a bit to sift through. (Protestant.)
# What's the more elegant solution? I imagine it would have the same rows, but with a summary of the columns (summed)

print(denom_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    denom = n()
  ))

ggplot(denom_summary) +
  geom_point(aes(relig, denom)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```
-->

## 15.4.1 Exercises

1.  **There are some suspiciously high numbers in `tvhours`. Is the mean a good
    summary?**

    No. Median would be a more realistic measure of central tendency.

2.  **For each factor in `gss_cat` identify whether the order of the levels is
    arbitrary or principled.**

```{r}
(columns <- which(lapply(gss_cat, class) == "factor"))

levels(gss_cat$marital) #arbitrary
levels(gss_cat$race) #arbitrary
levels(gss_cat$rincome) #principled
levels(gss_cat$partyid) #principled, politically right to left
levels(gss_cat$relig) #arbitrary
table(gss_cat$denom) #arbitrary
```


3.  **Why did moving "Not applicable" to the front of the levels move it to the
    bottom of the plot?**
    
    It's essentially a low value--position one in the list of levels.


## 15.5.1 Exercises

1.  **How have the proportions of people identifying as Democrat, Republican, and
    Independent changed over time?**
    
    A little shifting through the mid-aughts, but it looks like we're about back to where we were in 2000.
    
```{r}
  mutate(gss_cat, partyid = 
           fct_collapse(partyid,
                        Republican = c("Strong republican", "Not str republican"),
                        Democrat = c("Not str democrat", "Strong democrat"),
                        Independent = c("Ind,near rep", "Independent", "Ind,near dem"),
                        unnecessary = c("No answer", "Don't know", "Other party")
)) %>%
  filter(partyid != "unnecessary") %>%
  count(partyid, year) %>% # View()
  ggplot()+
  geom_col(aes(fct_rev(partyid), n)) +
  facet_wrap(~ year)
```
    
2.  **How could you collapse `rincome` into a small set of categories?**

```{r, fig.width=5, fig.height=3}
rincome_simple <- gss_cat

rincome_simple$rincome <- fct_lump(gss_cat$rincome) %>%
  fct_collapse(
    "Lt5000" = c("Lt $1000", "$1000 to 2999", "$3000 to 3999", "$4000 to 4999"),
    "5000to14999" = c("$5000 to 5999", "$6000 to 6999", "$7000 to 7999", "$8000 to 9999", "$10000 - 14999"),
    "15000ormore" = c("$15000 - 19999", "$20000 - 24999", "$25000 or more")
  )

filter(rincome_simple, rincome == "Lt5000" | rincome == "5000to14999" | rincome == "15000ormore") %>%
  ggplot()+
  geom_bar(aes(rincome))
```